Deep learning approaches on musical assisted-orchestration: an evaluation study

### 2020-05-11 paper/latex/paper.tex
Added to the paper

% Title.
% ------
\title{Neural models for computer-assisted musical orchestration: a preliminary study}

@@

%
\maketitle
%
\begin{abstract}
In this paper we will explore preliminary neural models for the task of computer-assisted musical orchestration. After an introduction on the problem, we will show how we decided to model musical orchestration as a classification task. In this context, we will propose two deep learning models and we will show how they perform as classifiers for musical instruments recognition by comparing them with specific baselines. We will the show how they perform, both qualitative and quantitative, in the task of computer-assisted orchestration by comparing them with state-of-the-art systems. We will highlight, finally, benefits and problems of neural approaches for orchestration and we will propose possible future steps.

\end{abstract}
%
\section{Introduction}\label{sec:introduction}

\section{Data generation}
We used the TinySOL database to create our input data. TinySOL is a subset of the Studio On Line (SOL) database created by IRCAM. TinySOL contains 2,478 samples from 12 instruments. The instruments come from different orchestral families: strings, woodwinds, and brass are all represented in TinySOL. Each TinySOL sample is one instrument playing one note in the "ordinary" playing style.

When training a model, the number of samples $n$ to be used in each combination is specified. To create the training data, $n$ instruments are randomly selected from the list of specified instruments, and then a TinySOL sample is randomly selected for each chosen instrument. This gives $n$ different samples, each of a different instrument. These samples are combined to form a soundfile that has each sample playing simultaneously. The audio was trimmed to four seconds in length so our data would all be of the same size. The melspectrogram is taken of this combination, creating the features to be fed to the model. We created the melspectrogram by \textbf{(Alejandro can you write this of how you generated the melspectrograms using the STFT?)}. We chose to use the melspectrogram because of its ability to represent timbre, a key part of assisted orchestration.

@@

Our idea is to perform orchestration using deep neural networks trained for classification. By generating input data that is a mixture of different instruments and pitches, the network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. When a target sound is then fed to the network, it attempts to classify which samples would best represent the target. By taking the samples that have the highest probability of being in the target, an orchestrated solution can be created.

We did not train our model to classify the dynamics of a sample despite TinySOL having pianissimo, mezzo forte, and fortissimo recordings for each sample. If we had, then each class would have only one data point. Instead, when the model is used for orchestration, the dynamic is determined by the probability of that sample as output by the model. If the model output a probability higher than $0.66$ for a sample, the fortissimo version of the sample was used. If it was between $0.33$ and $0.66$, then the mezzo forte version, and if less than $0.33$ the pianissimo dynamic.

\subsection{Baseline}
In order to have a baseline to compare our results against, we attempted to solve the classification problem using various parametric classifiers. The classifiers we tested were Support Vector Machine (SVM), Random Forest, and K-Nearest Neighbors. We used the implementations provided in the scikit-learn library for each classifier. For SVM, we used SVC with an RBF kernel. For Random Forest, we set the maximum depth of each tree to be 15. Each classifier was wrapped in a MultiOutputClassifier to achieve the multi-label nature of this problem. We found SVM to have the highest accuracy of the three classifiers across all experiments. All of the following baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample is a combination of one or more instruments and is four seconds in length. The features used are the mel frequency cepstral coefficients (MFCCs) of the resulting combination, with a total of 100 coefficients per sample. We chose MFCCs because of their ability to describe the timbre of a sample using a relatively small number of coefficients.

@@

For a slight modification on this setting, we no longer attempted to identify which instrument was present if it was not one of the two instruments whose pitch class was being identified. Instead, the classifier would simply identify that an instrument that was not one of the two was present. Since this is a simpler problem, it lead to increased accuracies. The accuracies from this experiment are outlined in Table 1.

@@

This was still a simplified version of the problem, as we were identifying only the pitch class of a few instruments. However, the parametric classifiers were unable to achieve accurate results as the number of instruments increased. Therefore, we did not attempt the full setting of the problem, in which individual pitches are classified for all instruments, with parametric classifiers.

\begin{table}
  \begin{center}
    \label{tab:table1}
    \begin{tabular}{|c|c|c|c|} 
    	  \hline
      \textbf{Instr. 1} & \textbf{Instr. 2} & \textbf{SVM Accuracy} & \textbf{RF Accuracy}\\
      \hline
      Violin & Flute & 38.8\% & 9.8\% \\
      \hline
      Violin & Trumpet & 33.8\% & 9.1\% \\
      \hline
      Violin & Cello & 34.8\% & 6.3\% \\
      \hline
      Cello & Viola & 32.1\% & 5.8\% \\
      \hline
      Oboe & French Horn & \textbf{39.9\%} & \textbf{17.5\%} \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Comparison of results between SVM and Random Forest. Each data point is a combination of two TinySOL samples, where at least one of the samples is from one of the two instruments specified for that experiment. For the samples drawn from one of the two instruments, the pitch class of that sample is identified. For a sample not from one of the two instruments, the classifier simply attempted to identify that a sample from one of the non-specified instruments was present. In this setting, there are 25 classes: 24 from the 12 pitch classes from 2 instruments, and 1 class that specified whether an instrument that is not one of the two specified is present.}
\end{table}

@@

The first deep model we tested was a Convolutional Neural Network (CNN). CNN shows good performance on classification assignment for its strong ability to extract spatial features, so we implemented a basic CNN with three convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a 2 * 2  MaxPool layer with a stride of 2. The kernel size is 3 * 3 with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32. This setting is simple but we can make a good comparison between it and the baseline.

- figure

Given that audio is one kind of sequential data, it is easy to consider Long Short-Term Memory network (LSTM) as test model. In our setting, we added one LSTM layer with 32 outputs units following three convolutional layers. The convolutional layers did not change the sequence relationship. Instead, it scaled down the size of feature maps. After LSTM layer, we reshaped the output into the same size before inputing. Then we implemented another convolutional layer with the same structure and kernel size. Finally, we flattened out the outputs and fed them into fully connected layers with Dropout layer. The probability of an element to be zeroed is 0.5 and we took sigmoid as activation function. We used binary classification for each class and treated them independently.

\subsection{ResNet}
- SVM was good but CNN was better and ResNet was best
- review all experiments we did and choose key plots

@@

@@

Although the data we trained the model on was fixed to four seconds of audio, the target samples can be of arbitrary length. By changing the mel hop length, which is the distance between the frames of the melspectrogram, an audio signal of any length can be represented as a matrix of the same size as the training data.

\subsection{Evaluation}
-evaluation of orchestrations (not evaluation of model)

We evaluated our orchestrations both qualitatively and quantitatively. Qualitative inspection was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution.

For quantitative evaluation, we orchestrated the same targets with OrchIdea, then compared the distance between our solutions and the targets and OrchIdea's solutions and the targets. The distance metric we use is defined in \eqref{distance}.

\begin{equation}\label{distance}
d(x, \tilde{x}) =\lambda_1 \sum_k \delta_{k1}(x_k - \tilde{x}_k) + \lambda_2 \sum_k \delta_{k2}|x_k - \tilde{x	}_k| \\
\end{equation}

$$
\delta_{k1} = 
\begin{cases}
1, \text{if   } x_k > \tilde{x}_k \\
0, \text{otherwise}
\end{cases} 
$$

$$
\delta_{k2} = 
\begin{cases}
1, \text{if   } x_k < \tilde{x}_k \\
0, \text{otherwise}
\end{cases} 
$$

This metric computes the difference between the FFT of the target and the FFT of the solution.

@@

\begin{table}
  \begin{center}
    \label{tab:table2}
    \begin{tabular}{|c|c|} 
    	  \hline
      \textbf{Model} & \textbf{Average Distance} \\
      \hline
      CNN with LSTM &   \\
      \hline
      ResNet &  \\
      \hline
      OrchIdea &  \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Comparison of the orchestration solutions of our two models, CNN with LSTM and ResNet, against OrchIdea's solutions. The distance between the target and solution are calculated for each target using \eqref{distance} and then averaged.}
\end{table}

@@

Based on our accuracies from training the model and the results of orchestrating 15 targets, our approach seems promising for orchestration. When comparing our method with OrchIdea, it is important to note that many of the processes OrchIdea uses to optimize its solution were not used in our model. This includes important aspects such as sparsity, partial filtering, \textbf{(Carmine any more to add?)}. Our model also lacks the implementation of symbolic constraints, which is an important part of assisted orchestration.

In order to have comparable solutions, we did not allow OrchIdea to create sparse solutions, meaning that OrchIdea was forced to use all 10 instruments in each solution. This creates a more fair comparison since our model is optimize solutions by using fewer than 10 samples.

\textbf{(Alejandro, add how there were certain instruments that performed really poorly or really well. Add how ResNet was awful for Flute and always output Flute C7)}

We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the pitch of the target. For a target that contained specific notes, the solution from ResNet contained both the note in the target and its partials: a sample that was two octaves higher, the fifth an octave up, and a minor third two octaves up (mix\_ObA4\_BnC3). Clearly, ResNet was able to identify the partials of the target and recreate them in the solution.

>>>>

% Title.
% ------
\title{Deep learning approaches on musical assisted-orchestration: an evaluation study}

@@

%
\maketitle
%
\begin{abstract}
The abstract should be placed at the top left column and should contain about 150-200 words.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}

\section{Data generation}

\subsection{Sources}
We used the TinySOL database to create our input data. This database contains samples of notes played using the ordinario style and with three different dynamics for each pair of instrument and pitch : \textit{pp}, \textit{mf} and \textit{ff}. The exact list of samples is shown in Tab.\ref{tab_pitch}.  \textbf{(Should we talk more about TinySOL here?)}

Since the goal was to train the model to detect multiple instruments playing simultaneously, the samples used for training were generated by mixing a random combination of instruments and pitches. More precisely, given the number $N$ of instruments playing at the same time, the data was generated as follows:

\begin{itemize}
\item An orchestra of $N$ instruments is chosen in advance (it is the same for all samples).
\item For each instrument in the orchestra, a random note is picked among the corresponding available samples in TinySOL.
\item For each chosen note, a random dynamic is picked.
\item The samples are added and normalized by the number of instruments.
\item The Mel spectrogram of the mixture is then computed and used as input of the model.
\end{itemize}

The choice of Mel spectrograms as input features of the model is very commmon in music information \textbf{\{Insert reference\}} and can be considered to be an accurate representation of the way humans perceive sound and music.\\

Samples had a sample rate of $44100Hz$ and were all padded or chopped to be exactly 4 seconds long. The Mel spectrograms were then generated using an FFT hop length of 2048 samples (e.g. the window of each FFT was $46ms$ wide), and a number of Mel bins equal to 128. Therefore, all the features fed to the model were matrices of size $128\times 345$. We used Librosa package in Python to compute the features, and more details on the exact computations can be found in \textbf{\{Reference to Librosa\}} The choice of the hop length was made by doing a compromise between the amount of information extracted by each FFT window, and the ability to capture changes in the dynamic, assuming that no change faster than $10ms$ could be perceived by the human ear.

\begin{table}
\begin{tabular}{|c|c|c|}
\hline
Instrument & Abbreviation & Range \\
\hline
Violin & Vn & G3-E7\\
Cello & Vc & C2-C6\\
Viola & Va & C3-C7\\
Trumpet C & TpC & F\#3-C6\\
Trombone & Tbn & A\#0;A\#1-C5\\
Oboe & Ob & A\#3-A6\\
Horn & Hn & G1-F5\\
Flute & Fl & B3-D7\\
Clarinette Bb & ClBb & D3-G6\\
Contrabass & Cb & E1-C5\\
Bass Tuba & BTb & F\#1-F4\\
Bassoon & Bn & A\#1-D\#5\\
\hline
\end{tabular}
\begin{caption}Table containing the pitch ranges in the TinySOL database, used for training. Hyphen means all the notes between the two explicited notes. \label{tab_pitch}
\end{caption}
\end{table}

\subsection{Data augmentation}

In order to increase the variability of the samples, we used two methods of data augmentation on the samples.\\

The first one was done on the mixture of the samples, by slightly shifting the frequency of the signal. In practice, the sample rate used to compute the Mel spectrogram was modified to be within $5\%$ of the acual $44100Hz$.\\

The second one was done on the Mel spectrogram, by setting to 0 random columns and rows of the feature matrices. For a given matrix, each column and each row had individually a $1\%$ chance to be set to 0, which yielded an average of $1.28$ reset columns and $3.45$ reset rows. This method of data augmentation aimed to be more resilient to the possible errors of recording, due either to a setup that induces interferences and cancels specific frequences, or to a default in the microphone that fails to record a small temporal portion.

@@

Our idea is to perform orchestration using deep neural networks trained for classification. By generating input data that is combinations of different instruments and pitches, the network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. Then, when a network is given a target sound as input, it will attempt to apply the same classification rules, outputing a vector of the probabilities of each sample being in the target sound. By taking the samples that have the highest probability, an orchestrated solution can be created.

We did not train our model to classify the dynamics of a sample despite TinySOL having pianissimo, mezzo forte, and fortissimo recordings for each sample. If we had, then each class would have only one data point. Instead, when the model is used for orchestration, the dynamic is determined by the probability of that sample as output by the model. If the model output a probability higher than $0.66$ for a sample, the fortissimo version of the sample was used. If it was between $0.33$ and $0.66$, then the mezzo forte version, and if less than $0.33$ the pianissimo dynamic. \textbf{\{I am not sure this hypothesis can be used accurately\}}

\subsection{Baseline}
In order to have a baseline to compare our results against, we attempted to solve the classification problem using various linear classifiers. The classifiers we tested were Support Vector Machine (SVM), Random Forest, and K-Nearest Neighbors. We used the implementations provided in the scikit-learn library for each classifier. For SVM, we used SVC with an RBF kernel. For Random Forest, we set the maximum depth of each tree to be 15. Each classifier was wrapped in a MultiOutputClassifier to achieve the multi-label nature of this problem. We found SVM to have the highest accuracy of the three classifiers across all experiments. All of the following baselin experiments used 50,000 generated samples with a train-test split of 60/40. Each sample is a combination of one or more instruments and is four seconds in length. The features used are the mel frequency cepstral coefficients (MFCCs) of the resulting combination, with a total of 100 coefficients per sample.

@@

For a slight modification on this setting, we no longer attempted to identify which instrument was present if it was not one of the two instruments whose pitch class was being identified. Instead, the classifier would simply identify that an instrument that was not one of the two was present. Since this is a simpler problem, it lead to increased accuracies. Flute and Violin went from 30.2\% to 38.8\%, Oboe and French Horn from 27.3\% to 39.9\%, and Trombone and Bass Tuba from 2.2\% to 3.5\%. \textit{(I think we should maybe put a table of this data instead of listing it all out)} Random Forest performed significantly worse for this setting. Flute and Violin had an accuracy of 9.8\% and Oboe and French Horn was 17.5\%. For this reason, we stopped testing with Random Forest and continued with SVM only.

@@

This was still a simplified version of the problem, as we were identifying only the pitch class of a few instruments. However, the linear classifiers were unable to achieve accurate results as the number of instruments increased. Therefore, we did not attempt the full setting of the problem, in which individual pitches are classified for all instruments, with linear classifiers.

@@

The first deep model we tested was a Convolutional Neural Network (CNN). CNN shows good performance on classification assignment for its strong ability to extract spacial features, so we implemented a basic CNN with three convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a 2 * 2  MaxPool layer with a stride of 2. The kernel size is 3 * 3 with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32. This setting is simple but we can make a good comparison between it and the baseline. \\

\begin{figure}
\includegraphics[scale=0.45]{CNN_LSTM_diagram.png}
\caption{Diagram of the CNN with LSTM architecture. \label{cnn_lstm}}
\end{figure}

Given that audio is one kind of sequential data, it is easy to consider Long Short-Term Memory network(LSTM) as test model. In our setting, we added one LSTM layer with 32 outputs units following three convolutional layers following the same pattern as the CNN described earlier. After LSTM layer, we reshaped the output into the same size as the output of the last convulational layer, which was fed to another convolutional layer with the same structure and kernel size, yielding a tensor of dimensions $32\times 8 \times 21$. Finally, we flattened out the outputs and fed them into fully a connected layer with Dropout layer, followed by another fully connected layer. Finally, the sigmoid function was applied to th final layer in order to get a probability between 0 and 1 for each individual label.\\

\subsection{ResNet}

The second deep model that we used was the well known deep residual network ResNet.

@@

\subsection{Classification results}

During training, the loss function used to optimize the inner parameters of the model was binary cross entropy, as it is the common choice for multiclass multilabel classification frameworks such as this one. However, the value of the loss function alone is difficult to interpret. This is why, we used a complementary function $f$ only for evaluation, to match the notion of accuracy of the model with the intuitive approach. This function compares a vector of expected outputs $\overline{X}$ with the estimated output from the model $\hat{X}$ by using the following function

\begin{equation}
f(\overline{X}, \hat{X}) = \frac{1}{N}<\overline{X}, M_N(\hat{X})>
\label{eval}
\end{equation}

where

\begin{equation}
M_N(\hat{X})_i = \left\{\begin{array}{ll}
1 \text{ if } i \in I_N(\hat{X})\\
0 \text{ otherwise}
\end{array}\right.
\label{NMax}
\end{equation}
and $I_N(\hat{X})$ is the set of the indices containing the $N$ first maximums of the vector $\hat{X}$.\\

More specifically, the function $M_N(\hat{X})$ takes as an input a vector of probabilities (values between 0 and 1) and outputs a vector where only the positions of the $N$ first maxima are set to 1. This new vector would be the orchestration of $N$ instruments given by the model. Thus, the function $f$ simply outputs the proportion of the estimated orchestration that matches the expected one.\\

Different experiments were made by varying the number of instruments in the orchestra $N$ but always matching this number with the number of instruments in each mixture sample. Tab. \ref{exp_tab} shows the orchestra for each different value of $N$ used in the experiments.\\

\begin{table}
\begin{tabular}{|c|l|}
\hline
$N$ & Orchestra\\
\hline
4 & Hn, Ob, Vn, Va\\
5 & Hn, Ob, Vn, Va, Vc\\
6 & Hn, Ob, Vn, Va, Vc, Fl\\
7 & Hn, Ob, Vn, Va, Vc, Fl, Tbn\\
8 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn\\
9 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn, TpC\\
10 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn, TpC, ClBb\\
\hline
\end{tabular}

\begin{caption}
Orchestra for each $N$ used in the experiments.
\label{exp_tab}
\end{caption}
\end{table}

Then, for both CNN with LSTM and ResNet, we computed the maximum accuracy over the epochs. Tab. \ref{epoch} compiles the number of samples per epoch and the total number of epochs used for each model and each value of $N$. Tab. \ref{class_res} shows the maximum accuracy computed using the function $f$ explicited in (\ref{eval}).

\begin{table}
\begin{tabular}{|c|c|c|c|}
\hline
Model & $N$ & Nb epochs & Nb samples\\
\hline
CNN-LSTM & 4-7 & 50 & 100000 \\
CNN-LSTM & 8-10 & 50 & 200000 \\
ResNet & 4-8 & 20 & 200000 \\
ResNet & 9-10 & 20 & 400000 \\
\hline
\end{tabular}

\begin{caption}
Number of epochs and number of samples per epoch used for training each model with a given value of the size of the orchestra $N$.
\label{epoch}
\end{caption}
\end{table}

\begin{table}
\begin{tabular}{|c|l|}
\hline
$N$ & Orchestra\\
\hline
4 & Hn, Ob, Vn, Va\\
5 & Hn, Ob, Vn, Va, Vc\\
6 & Hn, Ob, Vn, Va, Vc, Fl\\
7 & Hn, Ob, Vn, Va, Vc, Fl, Tbn\\
8 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn\\
9 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn, TpC\\
10 & Hn, Ob, Vn, Va, Vc, Fl, Tbn, Bn, TpC, ClBb\\
\hline
\end{tabular}

\begin{caption}
Orchestra for each $N$ used in the experiments.
\label{class_res}
\end{caption}
\end{table}

@@

We evaluated our orchestrations both qualitatively and quantitatively. Qualitative inspection was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that represented a C sharp minor chord. For a target that contained specific notes, the solution from ResNet contained both the note in the target and its partials: a sample that was two octaves higher, the fifth an octave up, and a minor third two octaves up (mix\_ObA4\_BnC3).

For quantitative evaluation, we orchestrated the same targets with OrchIdea, then compared the distance between our solutions and the targets and OrchIdea's solutions and the targets. The distance metric we used was

$$d(x^t, \tilde{x}^t) =\lambda_1 \sum_k \delta_k^t(x_k^t - \tilde{x}_k^t) + \sum_k \delta_k^t|x_k^t - \tilde{x	}_k^t|$$
$$\delta_k^t = 
\begin{cases}
1, \text{if} (x_k - \tilde{x}_k) < 0 \\
0, \text{otherwise}
\end{cases} $$

@@

@@

Based on our accuracies from training the model and the results of orchestrating 15 targets, our approach seems promising for orchestration. When comparing our method with OrchIdea, it is important to note that many of processes OrchIdea uses to optimize its solution were not used in our model. This includes important aspects such as sparsity, partial filtering, \textbf{(Carmine any more to add?)}. Our model also lacks the implementation of symbolic constraints, which is an important part of assisted orchestration.

\textbf{(Alejandro, add how there were certain instruments that performed poorly. Add how ResNet was awful for Flute and always output Flute C4)}

We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the pitch(es) of the target.
